A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2017-01-24 doc/paper1-submission/response.tex

% \newcommand{\benreplace}[2]{\benremove{#1} \benadd{#2}}

\newenvironment{reply}{$\triangleright$\bf}{$\triangleleft$}
\renewenvironment{quote}
               {\list{}{\rightmargin\leftmargin}%
                \item\relax\normalfont}
               {\endlist}
\sloppy
\begin{document}

\maketitle

Dear Dr. Wilke,

We appreciate the suggestions of Dr. Nakleh and the anonymous
reviewer. In response to their suggestions, we performed two
additional experiments on our pipeline to (1) ensure that the pipeline
could accurately recover a tree from its constituent pieces, and (2)
assessing the effect of input tree ranks on the supertree.

We also rewrote our section on notation and terminology, added a
discussion of other supertree methods, and clarified how our method
approaches tree weights differently than MRP and MinCut.

The first reviewer did not break their comments down into discrete
sub-points and so our reply is also in a somewhat discursive style.
However, we have responded to specific points as summarized by us.

@@

With these concerns out of the way I will add that I am uncertain whether a supertree method that achieves “perfect” (or nearly perfect) weighting can be devised. The only method I can think of that might allow principled estimation of tree weights would be a version of the maximum likelihood supertree method described by Steel and Rodrigo (2008). However, I doubt that such an ML supertree method would meet all of the desiderata for a good supertree method that R\&H articulate (especially their first goal). It would be valuable for the authors to expand on the potential advantages (which appear to be largely computational) and pitfalls of their extreme weighting method.

\begin{reply}
Response 2. We appreciate the thought that the reviewer put into articulating their concerns. The reviewer seems to assume that our supertree algorithm is designed to infer the true tree from the input trees like Bininda-Emonds.  However, we don't say that anywhere, and are not in fact trying to do that.  Instead we are trying to summarize the input trees, including low-quality inputs.  We agree with the reviewer that there are a lot of problems with trying to infer phylogenies from input trees, as expressed in the Gatesy et al (2004) critique of supertree methods.

We suspect a lot of readers may have the same question.  We therefore add the following text to the Background to clarify that we are not trying to infer the true tree from the input trees, and why not:
\begin{quote}
\textbf{Non-goal: phylogeny estimation}

Many researchers construct supertrees for the purpose of inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014}.  Such approaches treat input trees as data or as surrogates for data matrices \citep{gatesy2004critique}, and the supertree is seen as more accurate than the individual input trees because more input trees (and presumably more data) stand behind it.  For such methods, conflict resolution is a primary aim of the method.

Our supertree method operates in a different paradigm.  Phylogeny estimation is an explicit non-goal of our supertree algorithm.  We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method. Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.  Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.  Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.  The annotations file constructed by our pipeline is essential for this task because it includes information about conflict with our supertree as well as support for it.

Our method must resolve conflicts in order to construct a single supertree.  However, the rank information used to resolve conflicts is an input to the method, not an output from the method.  We thus perform curation-based conflict resolution, not inference-based conflict resolution.
\end{quote}

@@

With that clarification behind us, we can respond to a number of specific concerns:

2a. The reviewer seems concerned that low-quality source trees will undermine the estimation, and specifically that relationships present in more trees will dominate, leading to an incorrect estimate of the true tree.

However we are not seeking to estimate the true tree, and our rank-based approach to conflict resolution doesn't favor relationships that are more frequent:

@@

\begin{quote}
If an expert were quite confident in the c grouping, then she could assign that input node a high ranking. A supertree that used ranks could then recover this grouping even if its inclusion did not increase the total number of input nodes that are displayed by the summary tree.
\end{quote}

% We probably do need to state that, unlike MRP and MinCut, our approach doesn't favor relationships that occur more often.

2b. The reviewer seems concerned that our tree weights are extreme, and that it is difficult to estimate good tree weights from the input trees.

We appreciate these concerns and add the following text to explain our approach more clearly:

\begin{quote}
These large weights help to achieve our goal of allowing curators to fix problems in the tree.  While it is possible to view our pipeline as an optimization algorithm by assigning weights to the input trees in this way, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
\end{quote}

We sidestep the problem of estimating good tree weights (which is indeed difficult, if not impossible) by relying on curators to impose rankings instead.

@@

2c. We agree that it would be interesting to know how much effect the ranks had.  We therefore add the following section:
\begin{quote}
We ran the pipeline on the OTL4 data with the ranks reversed.  This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the \otlprop{}.  Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.  Thus, reversing the ranks makes a smaller different in the \otlprop{} supertree than the new synthesis algorithm.  We also note that a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.
\end{quote}

2d. We agree with the reviewer that applying ranks to trees rather than nodes is problematic, and have added the following to the discussion:
\begin{quote}
However, one downside of our current approach is that every node in an input
tree must recieve the same rank.  As a result, bootstrap proportions
and posterior probabilities are not taken into account.
\end{quote}

2e. We think that explaining the Hinchliff (2015) paper is beyond the scope of the current manuscript.
\end{reply}

@@

The authors do not address my third concern. The degree to which opentree will be used as a source of information for comparative methods is unclear to me. However, regardless of whether it is commonly used for that purpose there is an even more fundamental question: should opentree in its current releases be used by the for comparative methods by the community? I recognize that the R\&H methods could be used outside of opentree, but I think some readers may want an assessment of whether the tree is “ready for prime time.” Other than the broad coverage, is opentree as good as (or better than) than available synthetic trees (e.g., Bininda-Emonds et al. 2007 or Jetz et al. 2012)? I can think of at least one expression of concern regarding the use of taxonomies: the biogeographic study of Wang et al. (2016) that showed a major change in their conclusions when a taxon that appears to be misplaced by taxonomic constraints is included or excluded. The low weight of the taxonomy in opentree might ameliorate this issue. Obviously, addressing assessing this issue might be challenging (e.g., running the opentree analysis without the taxonomy and comparing that tree to a version of opentree pruned to exclude taxa that are only represented in the taxonomy). Although potentially useful I am not certain whether it is warranted.

\begin{reply}
Since we are not trying to construct a phylogeny estimate like Bininda-Edmonds, the question of whether our tree is ``ready for prime time'' may not be the right question.  However, since this may be the question of many readers, we add the following paragraph to the Conclusions to clarify that we are not claiming superior accuracy for our supertree:

\begin{quote}
Our supertree pipeline makes it possible to summarize and merge
information from across the Tree of Life.  However, we do not claim
that our supertree is more accurate than the input trees that it
summarizes.  Before using the output of our supertree pipeline
as input to an evolutionary analysis, researchers should first assess
the accuracy of the output supertree in the clade of interest,
possibly adjusting the ranks of the input trees to produce a more
useful outcome.  We have demonstrated that ranks supplied by curators
have substantial effects on the output of the pipeline.  This
can be considered to be a benefit of our approach, since it allows
curators to adjust the output of the pipeline, and it is easy to see
why the supertree contains the relationships that it does. However,
one downside of our current approach is that every node in an input
tree must recieve the same rank.  As a result, bootstrap proportions
and posterior probabilities are not taken into account.
\end{quote}
\end{reply}

@@

Wang N, Kimball RT, Braun EL, Liang B \& Zhang Z. 2016 (early online). Ancestral range reconstruction of Galliformes: the effects of topology and taxon sampling. Journal of Biogeography, doi: 10.1111/jbi.12782. http://onlinelibrary.wiley.com/doi/10.1111/jbi.12782/full

\newpage
\section*{Reviewer 2 (Luay Nakhleh)}
\subsection*{Basic reporting}
The authors report on a new method for constructing supertrees from an input set of trees. Supertree methods are essential if we're to construct trees with millions of taxa, which is what the authors pursue here. The authors describe four criteria that they argue a supertree construction method should satisfy. They then proceed to present a method that satisfies the first three (the method is a greedy algorithm, so it is not guaranteed to satisfy the fourth criterion). They ran the method on a data set of over 2 million taxa (in the resulting supertree) and compared the results to those obtained by the method of Hinchliff et al. (2015).
\subsection*{Experimental design}
The authors test their method on one data set and compare their findings to those obtained by the method of Hinchliff et al.
\subsection*{Validity of the findings}
The results reported on the data set, and comparisons to the method of Hinchliff et al. are reasonable.
\subsection*{Comments for the Author}
I have three general comments:

1. Related work: There is rich literature on methods for supertree methods. I suspect that these existing methods do not handle data sets with millions of taxa, which is why the authors probably chose not to discuss them. But, still, I'd encourage the authors to have a discussion (in the Background) on existing supertree methods and why they chose not to use them. In particular, it would be useful for the community to understand for existing methods which of the four criteria they satisfy.

\begin{reply}
We add the following two paragraphs to explain why we do not use MRP or MinCut, two commonly-used supertree methods.
\begin{quote}
\textbf{Comparison to other supertree methods}

Matrix Representation Parsimony (MRP) is one of the
most widely used supertree approaches \citep{baum1992combining,BinindaEmonds2007b,Davis2014}.
MRP attempts to encode branches of input trees as characters in a data
matrix, and then find an optimal tree according to the maximum
parsimony criterion. Such an approach does not provide a clear and transparent
explanation for including particular edges in the supertree.
MRP can also infer edges that exist in none of the input trees, which 
violates our Goal \#1 \citep{gatesy2004critique}.

Another commonly used method is MinCut \citep{semple2000supertree} or
Modified MinCut \citep{page2002modified}.  Like our approach, MinCut
is based on the BUILD algorithm \citep{AhoSSU1981}.  However, MinCut deals
with conflicts by modifying the BUILD algorithm to resolve
incompatibilities by discarding edges that are present in the smallest 
number of input trees.  This approach thus violates our Goal \#2 of
resolving conflict via ranks that can be altered by a curator to
influence the output tree.
\end{quote}
\end{reply}

@@

Also, a minor point with big implications: To prove a problem is NP-hard, a reduction *to* (not *from*) it is necessary. So, please replace "to Max-Clique" by "from Max-Clique" in Section 1.1.2.

\begin{reply}
Thanks - fixed.
\end{reply}

3. Performance: The authors chose one data set. Could they do an analysis of some additional data sets? For example, they can try the "perfect" case in which they take one of the supertrees they got, get out of it many subtrees (so, they know that these subtrees could be put back together into a supertree, and they know what that supertree is), and then run the methods?

\begin{reply}
Thanks for this suggestion.  We performed this experiment and added the following section describing the experiment and the results:
\begin{quote}
To test the reliability of our methods, we divided the \otlprop{} summary
tree into a collection of input trees. The input trees were obtained
by splitting the grafted tree at each taxonomy node, so that each
input tree has a taxonomy node at the root, and taxonomy nodes as
tips, but no taxonomy nodes internally.  Each leaf node that was an
internal taxon was then replaced with a leaf taxon. This approach
ensures that any taxa that were contested in the original analyses
remain contested in this new analysis.  We then constructed a summary
tree from these input trees using the propinquity pipeline.  The
resulting summary tree was identical to the \otlprop{} summary tree.
\end{quote}
\end{reply}

### 2017-01-24 doc/paper1-submission/paper5.tex
Update.

Cutting any edge \textbf{$j$ }of a rooted tree induces a bi-partition
$S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
$S_{2}(j)$. Such a bi-partition is called a rooted split and written
$S(j)=S_{1}(j)|\bullet S_{2}(j)$. Here $S_{1}(j)$ is called the ``cluster'' of $j$, and contains tip taxa on the side of edge $j$ that does not contain the
root. $S_{2}(j)$ contains the tip taxa on the side of edge $j$ that
contains the root. We refer to $S(j)$ as a \emph{rooted} split because
the right side of the split implicitly contains the root node, as indicated
by the $\bullet$.

@@

Our supertree method operates in a different paradigm.  Phylogeny estimation is an explicit non-goal of our supertree algorithm.  We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method. Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.  Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.  Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.  The annotations file constructed by our pipeline is an essential for this task because it includes information about conflict with our supertree as well as support for it.

@@

\subsection{Sensitivity of the \otlprop{} supertree to ranks}
We ran the pipeline on the OTL4 data with the ranks reversed.  This drastic alteration of the ranks lead to a summary supertree that conflicted with 1264 branches of the \otlprop{}.  Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.  Thus, reversing the ranks makes a smaller different in the \otlprop{} supertree than the new synthesis algorithm.  We also note that a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.

@@

Our supertree pipeline makes it possible to summarize and merge
information from across the tree of life.  However, we do not claim
that our supertree is more accurate than the input trees that is
summarizes.  We demonstrate that ranks supplied by curators have
substantial effects on the output of the pipeline.  Currently every
node in an input tree must recieve the same rank, and bootstrap
proportions and posterior probabilities are not taken into account.
Therefore, before using the output of our supertree pipeiline in an
evolutionary analysis, researchers must first assess the accuracy of
the output supertree, possibly adjusting the ranks of the input trees
to produce a more useful outcome.

>>>>

Cutting any edge $j$ of a rooted tree induces a bi-partition
$S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
$S_{2}(j)$. Such a bi-partition is called a rooted split and written
$S(j)=S_{1}(j)|\bullet S_{2}(j)$. Here $S_{1}(j)$ is called the ``cluster'' of $j$, and contains tip taxa on the side of edge $j$ that does not contain the
root. $S_{2}(j)$ contains the tip taxa on the side of edge $j$ that
contains the root. We refer to $S(j)$ as a \emph{rooted} split because
the right side of the split implicitly contains the root node, as indicated
by the $\bullet$.

@@

Our supertree method operates in a different paradigm.  Phylogeny estimation is an explicit non-goal of our supertree algorithm.  We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method. Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.  Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.  Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.  The annotations file constructed by our pipeline is essential for this task because it includes information about conflict with our supertree as well as support for it.

@@

\subsection{Sensitivity of the \otlprop{} supertree to ranks}
We ran the pipeline on the OTL4 data with the ranks reversed.  This drastic alteration of the ranks led to a summary supertree that conflicted with 1264 branches of the \otlprop{}.  Compared to this, the \otl{} supertree tree conflicts with 1387 branches of the \otlprop{} supertree.  Thus, reversing the ranks makes a smaller different in the \otlprop{} supertree than the new synthesis algorithm.  We also note that a total of 36533 edges of the grafted tree could be altered by changing the ranks, and most of these edges are unaffected.

@@

Our supertree pipeline makes it possible to summarize and merge
information from across the Tree of Life.  However, we do not claim
that our supertree is more accurate than the input trees that it
summarizes.  Before using the output of our supertree pipeline
as input to an evolutionary analysis, researchers should first assess
the accuracy of the output supertree in the clade of interest,
possibly adjusting the ranks of the input trees to produce a more
useful outcome.  We have demonstrated that ranks supplied by curators
have substantial effects on the output of the pipeline.  This
can be considered to be a benefit of our approach, since it allows
curators to adjust the output of the pipeline, and it is easy to see
why the supertree contains the relationships that it does.  However,
one downside of our current approach is that every node in an input
tree must recieve the same rank.  As a result, bootstrap proportions
and posterior probabilities are not taken into account.
